---
title: "mini_projects"
author: "Kai Movellan"
date: "10/26/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preparing the data


```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)
head(wisc.df)
```

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
```

```{r}
# Create diagnosis vector for later 
diagnosis <- wisc.df$diagnosis
diagnosis
```

## Exploratory data analysis

>Q1. How many observations are in this dataset?

```{r}
dim(wisc.data)
nrow(wisc.data)
#there are 569 observations in the data set
```

>Q2. How many of the observations have a malignant diagnosis?

```{r}
#use the table function to count the number of benign and malignant diagnosis
table (diagnosis)
#There are 212 malignant diagnosis
```

>Q3. How many variables/features in the data are suffixed with _mean?

```{r}
#Check the variables in the data 
colnames(wisc.df)
#Use grep to create a vector that has a count of the column names with "_mean"
#Use length to count the vector created by grep
length(grep("_mean",colnames(wisc.df)))
#there are 10 variables int eh data that are suffixed with "_mean"

```

## Performing PCA

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp( wisc.data, scale=TRUE )
wisc.pr
```

```{r}
# Look at summary of results
summary(wisc.pr)
```

>Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

- 0.4427 of the original variance is camptured by the first principal compoents (PC1)


>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

- 3 components describe at least 70% or the original variance in the data



>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

- 7 components describe at least 90% of the original variance in the data


## Interpreting PCA results

>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

- This plot is messy and difficult to interpret because it uses rownames as the plotting character for biplots 

```{r}
#Create a biplot of the wisc.pr using the biplot() function
biplot(wisc.pr)
```

```{r}
#Scatter plot observations by components 1 and 2
#we need to access the pca scores data using wisc.pr$x
#pc1 and pc2 are in [,1:2]
#use as.factor to make diagnosis a binary to differentiate colors 

plot(wisc.pr$x[,1:2], col=as.factor(diagnosis), xlab = "PC1", ylab = "PC2")
```

>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,3], col=as.factor(diagnosis), xlab = "PC1", ylab = "PC3")
```

- the component 2 explains more variance in the original data than component 3. The first plot has a cleaner seperation between the two subgroups.


```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, color=df$diagnosis) + 
  geom_point(alpha=0.5)
```

## Variance explained

```{r}
# Calculate variance of each component
#head(wisc.pr)
pr.var <- (wisc.pr$sdev)^2
head(pr.var)

```

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```


>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean",1]
```

>Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

-5 principles

```{r}
wisc.pr$center
```

##Hierarchical clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <-scale(wisc.data)
```

```{r}
data.dist <- dist(data.scaled)
```

```{r}
wisc.hclust <- hclust(data.dist, method="complete", members=NULL)
```

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

-height at 19

```{r}
plot(wisc.hclust)
abline(wisc.hclust, col="red", lty=2)
```

```{r}
#library(rgl)
#plot3d(wisc.pr$x[,1:3], xlab="PC 1", ylab="PC 2", zlab="PC 3", cex=1.5, size=1, type="s", col=grps)
```

## Selecting number of clusters

```{r}
#Use cutree() to cut the tree so that it has 4 clusters. Assign the output to the variable wisc.hclust.clusters
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
```

```{r}
#We can use the table() function to compare the cluster membership to the actual diagnoses.
table(wisc.hclust.clusters, diagnosis)
```

> Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

```{r}
#cluster 2
wisc.hclust.clusters1 <- cutree(wisc.hclust, k=2)
table(wisc.hclust.clusters1, diagnosis)

#cluster 10
wisc.hclust.clusters2 <- cutree(wisc.hclust, k=10)
table(wisc.hclust.clusters2, diagnosis)
```

```{r}
(357+164)/nrow(wisc.data)
(201+6)/nrow(wisc.data)
```


## Using different methods

>Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

- I like ward.D2 because it creates groups such that variance is minimized within clusters. It doesn't require squaring the Euclidean distances `dist()` are squared before inputing them to the `hclust()`

```{r}
wisc.hclustx <- hclust(data.dist, method="ward.D2", members=NULL)
plot(wisc.hclustx)
```


##  Combining methods

We take the results of our pca analysis and cluster in this space wisc.pr$x
```{r}
summary(wisc.pr)
wisc.pr$x[5,3]
```


```{r}
hclust(dist(wisc.pr$x[,1:3]))
```

```{r}
#Use the distance along the first 7 PCs for clustering 
wisc.pr.hclust <- hclust(data.dist, method="ward.D2", members=NULL)
```


```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

crops table compare of diagnosis and my cluster groups

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```


```{r}
table (diagnosis, grps)
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
g <- as.factor(grps)
levels(g)
```

```{r}
g <- relevel(g,2)
levels(g)
```


```{r}
# Plot using our re-ordered factor 
plot(wisc.pr$x[,1:2], col=g)
```


```{r}
#install.packages("rgl")
library(rgl)
plot3d(wisc.pr$x[,1:3], xlab="PC 1", ylab="PC 2", zlab="PC 3", cex=1.5, size=1, type="s", col=grps)
```


## Sensitivity/ Specificity

```{r}
#Use the distance along the first 7 PCs for clustering 
wisc.pr.hclust <- hclust(data.dist, method="ward.D2", members=NULL)
```


```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

>Q15. How well does the newly created model with four clusters separate out the two diagnoses?

-the new model has 90.86% accuracy

```{r}
# Compare to actual diagnoses
table(wisc.pr.hclust.clusters, diagnosis)
```

>Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

-using wisc.km$cluster K-means clustering works the best in seperating diagnosis

```{r}
table(wisc.hclust.clusters, diagnosis)
```


**Accuracy** what proportion did we get correct if we call cluster 1 M and cluster 2B

```{r}
(329+188)/nrow(wisc.data)
(175+343)/nrow(wisc.data)
(165+343)/nrow(wisc.data)
```

**Sensitivity**
```{r}
(188)/(188+28)
(175)/(175+14)
(165)/(165+12)
```

**Specificity**
```{r}
329/(329+24)
343/(343+37)
343/(343+40)
```

>Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

-specifcity: wisc.pr.hclust.clusters
-sensitivity: wisc.hclust.clusters

## Prediction

Here we read some new data and use PCA model to examine whether they most closely resemble M or B patients from our origional dataset

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

Plot onto our PCA model 
```{r}
plot(wisc.pr$x[,1:2], col=as.factor(diagnosis))
points(npc[,1],npc[,2], col="green", pch=16, cex=3)
text(npc[,1],npc[,2], labels = c(1,2), col="blue")
```

>Q18. Which of these new patients should we prioritize for follow up based on your results?

-patient 2 id more malignant 


