---
title: "class17"
author: "Kai Movellan"
date: "11/26/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

>Q1. What column details the total number of people fully vaccinated? 

```{r}
colnames(vax)
```
9

>Q2. What column details the Zip code tabulation area? 

```{r}
colnames(vax)
```
2

>Q3. What is the earliest date in this dataset? 

```{r}
head(sort(vax$as_of_date,decreasing = TRUE))
```

2021-01-05

>Q4. What is the latest date in this dataset? 

```{r}
head(sort(vax$as_of_date,decreasing = FALSE))
```

2021-11-23

```{r}
skimr::skim(vax)
```

>Q5. How many numeric columns are in this dataset? 

9

>Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

8256

>Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
signif(((8256/81144)*100),2)
```


>Q8. [Optional]: Why might this data be missing?

```{r}
sum( is.na(vax$persons_fully_vaccinated) )
```

these people could be too young to get vaccinated or just not be getting vaccinated

##Working with dates

```{r warning=FALSE}
library(lubridate)
```
What is today’s date (at the time I am writing this obviously)
```{r}
today()
```


```{r}
# This will give an Error!
#today() - vax$as_of_date[1]
```

```{r}
# Speciffy that we are using the Year-mont-day format
vax$as_of_date <- ymd(vax$as_of_date)
```


How many days have passed since the first vaccination reported in this dataset?
```{r}
today() - vax$as_of_date[1]
```

How many days have passed since the first vaccination reported in this dataset?
```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

>Q9. How many days have passed since the last update of the dataset? 

```{r}
today() - vax$as_of_date[nrow(vax)]
```

6

>Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
length(unique(vax$as_of_date))
```

46

##Working with ZIP codes

```{r}
#install.packages("zipcodeR")
library(zipcodeR)
```


```{r}
geocode_zip('92037')
```

Calculate the distance between the centroids of any two ZIP codes in miles, e.g.
```{r}
zip_distance('92037','92109')
```

More usefully, we can pull census data about ZIP code areas (including median household income etc.)
```{r}
reverse_zipcode(c('92037', "92109") )
```

```{r}
# Pull data for all ZIP codes in the dataset
#zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

##Focus on the San Diego area

```{r}
# Subset to San Diego county only areas
sd <- vax[vax$county == "San Diego" , ]
```

Using dplyr the code would look like this:
```{r warning=FALSE}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

Using dplyr is often more convenient when we are subsetting across multiple criteria - for example all San Diego county areas with a population of over 10,000.
```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
head(sd.10)
length(sd.10)
```

>Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
length(unique(sd))
```


>Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}

which.max(sd$age12_plus_population) 
```

>Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”? 

```{r}
sd.11 <- filter(sd, as_of_date == "2021-11-09")
vacinationsandiego<-(na.omit(sd.11$percent_of_population_fully_vaccinated))
mean(vacinationsandiego)
```


>Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?

```{r}
library(ggplot2)

hist(vacinationsandiego)
```

##Focus on UCSD/La Jolla

UC San Diego resides in the 92037 ZIP code area and is listed with an age 5+ population size of 36,144.
```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population

```

>Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) +
  aes(ucsd$as_of_date,
      ucsd$percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

##Comparing 92037 to other similar sized areas?

Let’s return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on as_of_date “2021-11-16”.
```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")

head(vax.36)
```

>Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
sd.mean<-mean(vax.36$percent_of_population_fully_vaccinated)
```

```{r}
ggplot(ucsd) +
  aes(ucsd$as_of_date,
      ucsd$percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")+
  geom_hline(aes(yintercept=0.6640413),color="red", lty=2)
```

>Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

>Q18. Using ggplot generate a histogram of this data.

```{r}
library(ggplot2)
p<-ggplot(vax.36, aes(x=vax.36$percent_of_population_fully_vaccinated)) + 
  geom_histogram(color="black") +
  labs(x="Percent Vaccinated")
p
```

>Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?


```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```
above

>Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36000)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(c(0,1)) +
  labs(x="Date", y= "Percent Vaccinated",
       title="Vaccination Rate Across California",
       subtitle="Only Areas With a Population Above 36k are Shown") +
  geom_hline(yintercept=0.6640413 , linetype=2)
```

>Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?

I am vaccinated and tested negative for covid so I'm excited to see my family