---
title: "halloweenminiproject"
author: "Kai Movellan"
date: "10/28/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##1. Importing candy data

```{r}
#use read.csv to import the "candy-data.csv" fi;e
candy_file <- "candy-data.csv"

candy <-read.csv(candy_file, row.names=1)
head(candy, n=7)
```
>Q1. How many different candy types are in this dataset?

- 85 types of candy

```{r}
#the rows show each type of candy in the dataset
dim(candy)
nrow(candy)
```

>Q2. How many fruity candy types are in the dataset?

- 38 fruity candies

```{r}
#use sum to add up all the binary 1s in the fruity column 
fruity<-candy$fruity
fruity
sum(fruity)
```

##2. What is your favorate candy?
```{r}
candy["Twix", ]$winpercent
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

- Twizzlers: 45.46628

```{r}
candy[,0]
candy["Twizzlers", ]$winpercent
```


>Q4. What is the winpercent value for “Kit Kat”?

- Kit Kat: 76.7686

```{r}
candy["Kit Kat", ]$winpercent
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

- Tootsie Roll Snack Bars: 49.6535

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
#install.packages("skimr")
library("skimr")
skim(candy)
```
>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

- winpercent is in a different scale 


>Q7. What do you think a zero and one represent for the candy$chocolate column?


```{r}
candy$chocolate
```

>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```


>Q9. Is the distribution of winpercent values symmetrical?

-no

```{r}
dist(candy$winpercent)
```

>Q10. Is the center of the distribution above or below 50%?

-below
```{r}
mean(candy$winpercent)
median(candy$winpercent)
```


>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

-chocolate is higher ranked than fruity candy

```{r}
# first we need to find all the chocolate candy rows in the `candy` data
choco<-as.logical(candy$chocolate)
#candy[choco,]$winpercent
choco.2<-candy[choco,]$winpercent
mean(choco.2)#=60.92153


fty<-as.logical(candy$fruity)
#candy[fty,]$winpercent
fty.2<-candy[fty,]$winpercent
mean(fty.2)#=44.11974

ngt<-as.logical(candy$nougat)
#candy[ngt,]$winpercent
ngt.2<-candy[ngt,]$winpercent
mean(ngt.2)#=60.05188

```


>Q12. Is this difference statistically significant?

-yes

```{r}
t.test(choco.2,fty.2)
```

>Q13. What are the five least liked candy types in this set?

- Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbuster

```{r}
head(candy[order(candy$winpercent),], n=5)
```


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
#install.packages("dplyr")
#library(dplyr)
#candy %>% arrange(winpercent) %>% head(5)
head(sort((candy$winpercent), decreasing=TRUE, n=5))
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col(alpha=0.5)
  
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
#improve this plot by recording by the `winpercent` value
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
  
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 

```

>Q17. What is the worst ranked chocolate candy?

-sixlet

>Q18. What is the best ranked fruity candy?

-nik l nip

##4. Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
my_cols[as.logical(candy$fruity)]="red"

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=2, max.overlaps = 8)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

-twix bars

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

-

## 5 Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

-chocolate and fruity

Q23. Similarly, what two variables are most positively correlated?
- bar and chocolate
-win percent and chocolate
-chocolate and pricepercent
-nougat and bar

##6. Principal Component Analysis

```{r}
pca <- prcomp(candy,scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16) +
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```


```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```


```{r}
library(ggrepel)

p + geom_text_repel(size=2, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#install.packages("plotly")
library(plotly)

ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

fruity, hard, and pluribus
these make snese 